Component response rate variation underlies the stability of complex systems (revision notes)

### 2020-01-23 notebook/revision_notes.Rmd
Some progress, finally get the eigenvalue spectra variance correct

@@

Variation in $\gamma$ ($\sigma^{2}_{\gamma}$) **can increase or decrease the radius of the circle**.

<!--- still wrong below -- no simple analytical solution seems possible --->

$$2 \sigma_{\gamma} \sigma_{\textbf{A}} \sqrt{SC}.$$

@@

For poisson, it's $$\sigma^{2}_{\gamma} \sigma_{\textbf{A}} \sqrt{SC}.$$

The mean of a uniform is $(1/2)b$, while the mean of the Poisson is $\lambda = \sigma^{2}_{\gamma}$

Note that the error around the expected radius of eigenvalues is not normally distributed. It is log distributed (check).

>>>>

```{r, echo = FALSE}
source("../R/utilities.R");
source("../R/random_mats.R");
```

@@

Introducing variation in $\gamma$ increases the total variation in the system, making it more likely that $\textbf{M} = \gamma\textbf{A}$ is unstable. Variation of the off-diagonal elements in $\textbf{M}$ is described by the joint variation of two random variables (to simplify the notation, $\sigma^{2}_{M}$ and $\sigma^{2}_{A}$ refer to variances of off-diagonal elements only),

$$\sigma^{2}_{M} = \sigma^{2}_{A}\sigma^{2}_{\gamma} + \sigma^{2}_{A}E[\gamma_{i}]^{2}+\sigma^{2}_{\gamma}E[A_{ij}]^{2}.$$

Note that in my simulations $E[\gamma_{i}] = 1$ and $E[A_{ij}] = 0$, so the above can be simplified,

$$\sigma^{2}_{M} = \sigma^{2}_{A}(1 + \sigma^{2}_{\gamma}).$$

The increase caused by $\sigma^{2}_\gamma$ can be visualised from the eigenvalue spectra of $\textbf{A}$ versus $\textbf{M} = \gamma\textbf{A}$. Given $d = 0$ and $C = 1$, the distribution of eigenvalues of $\textbf{M}$ and $\textbf{A}$ lie within a circle of a radius $\sigma^{2}_{M}\sqrt{S}$ and $\sigma^{2}_{A}\sqrt{S}$, respectively.

```{r, fig.width = 7, fig.height = 7, echo = FALSE}
S     <- 400;
C     <- 1;
dval  <- 0;
sigma <- 1/(1*sqrt(S));
a     <- 0;
b     <- 2;
mn    <- 0;

A_dat  <- rnorm(n = S * S, mean = mn, sd = sigma);
A_mat  <- matrix(data = A_dat, nrow = S);
C_dat  <- rbinom(n = S * S, size = 1, prob = C);
C_mat  <- matrix(data = C_dat, nrow = S, ncol = S);
A_mat  <- A_mat * C_mat;
gammas <- runif(n = S, min = a, max = b);
vargam <- (1/12) * (b-a)^2;

diag(A_mat) <- dval;
A1     <- gammas * A_mat;
A0     <- A_mat;
A0     <- A0 * mean(gammas);
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

joint_var <- ((sigma^2) * vargam) + ((sigma^2) * mean(gammas)^2) + 
             (vargam * (dval*mn)^2);

par(mfrow = c(2, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0.2, 0.2));
plot(A0_r, A0_i, pch = 4, cex = 0.7, xlab = "", ylab = "", cex.lab = 1.3, 
     cex.axis = 1.25, asp = 1, col = "dodgerblue4", xlim = c(-3, 1.5),
     xaxt = "n", ylim = c(-1.5, 1.5));
vl        <- seq(from = 0, to = 2*pi, by = 0.001);
A0x1a     <- mean(gammas) * sigma * sqrt(S) * cos(vl) + mean(diag(A0));
A0y1a     <- mean(gammas) * sigma * sqrt(S) * sin(vl);
points(x = A0x1a, y = A0y1a, type = "l", lwd = 3, col = "dodgerblue4");
text(x = -2.75, y = 1.75, labels = "a", cex = 2);

plot(A1_r, A1_i, pch = 4, cex = 0.7, xlab = "", ylab = "", cex.lab = 1.3, 
     cex.axis = 1.25, asp = 1, col = "firebrick", xlim = c(-3, 1.5),
     xaxt = "n", yaxt = "n", ylim = c(-1.5, 1.5));
crad      <- sqrt(joint_var) * sqrt(S);
A1x1a     <- crad * cos(vl) + mean(diag(A1));
A1y1a     <- crad * sin(vl);
points(x = A1x1a, y = A1y1a, type = "l", lwd = 3, col = "firebrick");
text(x = -2.75, y = 1.75, labels = "b", cex = 2);

calc_var <- A1;
diag(calc_var) <- NA;
calc_var  <- as.numeric(calc_var);
cvar1 <- var(calc_var[!is.na(calc_var)]);

################################################################################
################################################################################

S     <- 400;
C     <- 1;
dval  <- -1;
sigma <- 1/(1*sqrt(S));
a     <- 0;
b     <- 2;
mn    <- 0;

A_dat  <- rnorm(n = S * S, mean = mn, sd = sigma);
A_mat  <- matrix(data = A_dat, nrow = S);
C_dat  <- rbinom(n = S * S, size = 1, prob = C);
C_mat  <- matrix(data = C_dat, nrow = S, ncol = S);
A_mat  <- A_mat * C_mat;
gammas <- runif(n = S, min = a, max = b);
vargam <- (1/12) * (b-a)^2;

diag(A_mat) <- dval;
A1     <- gammas * A_mat;
A0     <- A_mat;
A0     <- A0 * mean(gammas);
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

joint_var <- ((sigma^2) * vargam) + ((sigma^2) * mean(gammas)^2) + 
             (vargam * (dval*mn)^2);

plot(A0_r, A0_i, pch = 4, cex = 0.7, xlab = "", ylab = "", cex.lab = 1.3, 
     cex.axis = 1.25, asp = 1, col = "dodgerblue4", ylim = c(-1.5, 1.5),
     xlim = c(-3, 1.5));
A0x1a     <- mean(gammas) * sigma * sqrt(S) * cos(vl) + mean(diag(A0));
A0y1a     <- mean(gammas) * sigma * sqrt(S) * sin(vl);
points(x = A0x1a, y = A0y1a, type = "l", lwd = 3, col = "dodgerblue4");
text(x = -2.75, y = 1.75, labels = "c", cex = 2);

plot(A1_r, A1_i, pch = 4, cex = 0.7, xlab = "", ylab = "", cex.lab = 1.3, 
     cex.axis = 1.25, asp = 1, col = "firebrick", yaxt = "n", 
     xlim = c(-3, 1.5), ylim = c(-1.5, 1.5));
text(x = -2.75, y = 1.75, labels = "d", cex = 2);

calc_var <- A1;
diag(calc_var) <- NA;
calc_var  <- as.numeric(calc_var);
cvar2 <- var(calc_var[!is.na(calc_var)]);

mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

The effect of $\gamma$ does not necessarily result in a circular eigenvalue spectra of $\textbf{M}$. In both panels of Figure 1, $S =$ `r S` and $\sigma_{A} = 1/\sqrt{S}$. In Figures 1a,b and 1c,d, $d = 0$ and $d = 1$, respectively, causing the eigenvalue spectra of $\textbf{A}$ to differ on the real axis (Figure a,c). Figure panels b and c show eigenvalue spectra after $\gamma$ is included such that $M = \gamma A$, where elements of $\gamma$ are drawn from a uniform distribution $\gamma \sim \mathcal{U}(0, 2)$. In both cases, $\sigma^{2}_{A} =$ `r round(sigma, digits = 4)` and $\sigma^{2}_{\gamma} =$ `r round(joint_var, digits = 4)`. Nevertheless, the change in eigenvalue spectra caused by $\gamma$ can affect $\max(\Re(\lambda))$ and therefore system stability.

Stability also decreases with increasing $\rho$ such that $\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right)$. Hence, when $\rho$ values are negative, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability is expected to increase [@Allesina2012]. When elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the covariance between $M_{ij}$ and $M_{ji}$,

$$\rho = \frac{cov(M_{ij}, M_{ji})}{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})}.$$ 
Because $\gamma$ does not affect $cov(M_{ij}, M_{ji})$, the sign of $\rho$ remains unchanged.

More down
======================

@@

### 2020-01-23 notebook/revision_notes.tex
Some progress, finally get the eigenvalue spectra variance correct

@@

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-4-1.pdf}

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\(\times 10^{-6}\)
(95\% CIs: -73.6979659\(\times 10^{-6}\), 83.9237307\(\times 10^{-6}\)).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-8-1.pdf}

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-9-1.pdf}

\section{Trying to think
analytically}\label{trying-to-think-analytically}

Trying to think analytically, note that the correlation \(\rho\) adjusts
the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{1}},

@@

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\). Allesina and Tang note that \(\rho\) is
defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{1}},

@@

Gibbs et al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{2}}
showed analytically that as \(S \to \infty\), the effect of a diagonal
matrix \(\textbf{X}\) on the stability of \(\textbf{M} = \textbf{XA}\)
becomes negligible. They also note that local stability of a random
matrix can be predicted from simple moments affecting three parameters,

@@

@@

\hypertarget{refs}{}
\hypertarget{ref-Allesina2015a}{}
1. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

\hypertarget{ref-Gibbs2017}{}
2. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).

>>>>

\section{Effects of correlations
interactions}\label{effects-of-correlations-interactions}

If the off-diagonal elements of \(\textbf{A}\) are sampled independently
from an identical distribution, then \(\max(\Re(\lambda))\) for
\(\textbf{M} = \textbf{A}\) can be estimated from five
values\textsuperscript{\protect\hyperlink{ref-Tang2014b}{1}}. These
values include (1) system size (\(S\)), (2) mean self-regulation of
components (\(d\)), (3) mean interaction strength between components
(\(\mu\)), (4) the standard deviation between component interaction
strengths (\(\sigma\)), and (5) the correlation of interaction strengths
between components, \(M_{ij}\) and \(M_{ji}\) (\(\rho\)). When
investigating the effect of varying component response rate
\(Var(\gamma)\) on stability by defining
\(\textbf{M} = \gamma\textbf{A}\), \(S\) remains unchanged. Further,
values of \(\gamma_{i}\) were sampled such that \(E[d]\) and \(E[\mu]\)
also remained unchanged (in practice, diagonal elements of
\(\textbf{M}\) were standardised so that mean values were identical
before and after adding \(\gamma\)). What \(Var(\gamma)\) does change is
the variation in component interaction strengths, and \(\rho\).

Introducing variation in \(\gamma\) increases the total variation in the
system, making it more likely that \(\textbf{M} = \gamma\textbf{A}\) is
unstable. Variation of the off-diagonal elements in \(\textbf{M}\) is
described by the joint variation of two random variables (to simplify
the notation, \(\sigma^{2}_{M}\) and \(\sigma^{2}_{A}\) refer to
variances of off-diagonal elements only),

\[\sigma^{2}_{M} = \sigma^{2}_{A}\sigma^{2}_{\gamma} + \sigma^{2}_{A}E[\gamma_{i}]^{2}+\sigma^{2}_{\gamma}E[A_{ij}]^{2}.\]

Note that in my simulations \(E[\gamma_{i}] = 1\) and \(E[A_{ij}] = 0\),
so the above can be simplified,

\[\sigma^{2}_{M} = \sigma^{2}_{A}(1 + \sigma^{2}_{\gamma}).\]

The increase caused by \(\sigma^{2}_\gamma\) can be visualised from the
eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \gamma\textbf{A}\). Given \(d = 0\) and \(C = 1\), the
distribution of eigenvalues of \(\textbf{M}\) and \(\textbf{A}\) lie
within a circle of a radius \(\sigma^{2}_{M}\sqrt{S}\) and
\(\sigma^{2}_{A}\sqrt{S}\), respectively.

@@

The effect of \(\gamma\) does not necessarily result in a circular
eigenvalue spectra of \(\textbf{M}\). In both panels of Figure 1,
\(S =\) 400 and \(\sigma_{A} = 1/\sqrt{S}\). In Figures 1a and 1b,
\(d = 0\) and \(d = 1\), respectively, causing the eigen spectra of
\(\textbf{A}\) to differ on the real axis (blue points). Red points she
eigen spectra after \(\gamma\) is include such that \(M = \gamma A\),
where elements of \(\gamma\) are drawn from a uniform distribution
\(\gamma \sim \mathcal{U}(0, 2)\). In both cases, \(\sigma^{2}_{A} =\)
0.05 and \(\sigma^{2}_{\gamma} =\) 0.0033.

\section{More down}\label{more-down}

The figure below shows the relationship between \(\sigma_{\gamma}\) and
the radius of the circle in which eigenvalue spectra are contained given
\(S = 1024\), \(C = 1\), and \(\sigma_{\textbf{A}} = 1/\sqrt{S}\).
Values of \(\gamma\) were drawn from a uniform distribution
\(\gamma \sim \mathcal{U}(0, b)\) at increasing values of \(b\) so that
\(\sigma_{\gamma} = (1/12)b^2\) gradually increased.

Note that the correlation \(\rho\) adjusts the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{2}},

@@

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-3-1.pdf}

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\). Allesina and Tang note that \(\rho\) is
defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{2}},

@@

Gibbs et al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{3}}
showed analytically that as \(S \to \infty\), the effect of a diagonal
matrix \(\textbf{X}\) on the stability of \(\textbf{M} = \textbf{XA}\)
becomes negligible. They also note that local stability of a random
matrix can be predicted from simple moments affecting three parameters,

@@

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-5-1.pdf}

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\(\times 10^{-6}\)
(95\% CIs: -74.3396367\(\times 10^{-6}\), 77.6796914\(\times 10^{-6}\)).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-12-1.pdf}

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-13-1.pdf}

@@

\hypertarget{refs}{}
\hypertarget{ref-Tang2014b}{}
1. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\hypertarget{ref-Allesina2015a}{}
2. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

\hypertarget{ref-Gibbs2017}{}
3. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).
